Get Your Own Just-In-Time Specializing Compiler For Free

### 2007-05-29 talk/dls2007/paper.tex
Finish LaTeXification, complete things a bit.

One of the central goals of the PyPy project is to automatically
produce dynamic compilers from an interpreter, with as little
modifications of the interpreter code base itself as possible.

The forest of flow graphs that the translation process \cite{VMCDLS}
generates and transforms constitutes a reasonable base for the
necessary analyses.  That's a further reason why having a high-level
runnable and analyzable interpreter implementation was always a
central tenet of the project: in our approach,
the dynamic compiler is just another aspect
transparently introduced by and during the translation
process.

Partial evaluation techniques should, at least theoretically,
allow such a derivation of a compiler from an interpreter [PE], but it
is not reasonable to expect the code produced for an input program by
a compiler derived using partial evaluation to be very good,
especially in the case of a dynamic language.  Essentially, the input
program doesn't contain enough information to generate good code; for
example the input program contains mostly no kind of type
information in that case.

@@

Inspired by Psyco \cite{PSYCO}, which is a hand-written dynamic compiler
based on partial evaluation for Python, we developed a technique -
*promotion* - for our dynamic compiler generator. Simply put, promotion
on a value stops compilation and waits until the run-time reaches this
point.  When it does, the actual run-time value is promoted into a
compile-time constant, and compilation resumes with this extra
information.

@@

Partial evaluation is the process of evaluating a function, say ``f(x,
y)``, with only partial information about the values of its arguments,
say the value of the ``x`` argument only.  This produces a *residual*
function ``g(y)``, which takes less arguments than the original -- only
the information not specified during the partial evaluation process needs
to be provided to the residual function, in this example the ``y``
argument.

Partial evaluation (PE) comes in two flavors:

* *On-line* PE: a compiler-like algorithm takes the source code of the
  function ``f(x, y)`` (or its intermediate representation, i.e. its
  control flow graph in PyPy's terminology), and some partial
  information, e.g. ``x=5``.  From this, it produces the residual
  function ``g(y)`` directly, by following in which operations the
  knowledge ``x=5`` can be used, which loops can be unrolled, etc.

* *Off-line* PE: in many cases, the goal of partial evaluation is to
  improve performance in a specific application.  Assume that we have a
  single known function ``f(x, y)`` in which we think that the value of
  ``x`` will change slowly during the execution of our program -- much
  more slowly than the value of ``y``.  An obvious example is a loop
  that calls ``f(x, y)`` many times with always the same value ``x``.
  We could then use an on-line partial evaluator to produce a ``g(y)``
  for each new value of ``x``.  In practice, the overhead of the partial
  evaluator might be too large for it to be executed at run-time.
  However, if we know the function ``f`` in advance, and if we know
  *which* arguments are the ones that we will want to partially evaluate
  ``f`` with, then we do not need a full compiler-like analysis of ``f``
  every time the value of ``x`` changes.  We can precompute once and for
  all a specialized function ``f1(x)``, which when called produces the
  residual function ``g(y)`` corresponding to ``x``.  This is *off-line
  partial evaluation;* the specialized function ``f1(x)`` is called a
  *generating extension*.

The PyPy JIT generation framework is based on off-line partial
evaluation.  The function called ``f(x, y)`` above is typically the main
loop of some interpreter written in RPython.  The size of the interpreter can range
from a three-liner used for testing purposes to the whole of PyPy's
Python interpreter.  In all cases, ``x`` stands for the input program
(the bytecode to interpret) and ``y`` stands for the input data (like a
frame object with the binding of the input arguments and local
variables).  Our framework is capable of automatically producing the
corresponding generating extension ``f1(x)``, which takes an input
program only and produces a residual function ``g(y)``.  This ``f1(x)``
is a compiler\footnote{
    What we get in PyPy is more precisely a \emph{just-in-time compiler:}
    if promotion is used, compiling ahead of time is not possible.
}
for the very same language for which ``f(x, y)`` is
an interpreter.

Off-line partial evaluation is based on *binding-time analysis,* which
is the process of determining among the variables used in a function (or
a set of functions) which ones are going to be known in advance and
which ones are not.  In the example of ``f(x, y)``, such an analysis
would be able to infer that the constantness of the argument ``x``
implies the constantness of many intermediate values used in the
function.  The *binding time* of a variable determines how early the
value of the variable will be known.

Once binding times have been determined, one possible approach to
producing the generating extension itself is by self-applying on-line
partial evaluators.  This is known as the second Futamura projection
\cite{FU}.  So far it is unclear if this approach can lead to optimal
results, or even if it scales well.  In PyPy we selected a more direct
approach: the generating extension is produced by transformation of the
control flow graphs of the interpreter, guided by the binding times.  We
call this process *timeshifting*.

@@

PyPy contains a framework for generating just-in-time compilers using
off-line partial evaluation.  As such, there are three distinct phases:

* *Translation time:* during the normal translation of an RPython
  program, say PyPy's Python interpreter, we perform binding-time
  analysis and off-line specialization ("timeshifting") of the
  interpreter.  This produces a generating extension, which is linked
  with the rest of the program.

* *Compile time:* during the execution of the program, when a new
  bytecode is about to be interpreted, the generating extension is
  invoked instead.  As the generating extension is a compiler, all the
  computations it performs are called compile-time computations.  Its
  sole effect is to produce residual code.

* *Run time:* the normal execution of the program (which includes the
  time spent running the residual code created by the generating
  extension).

@@

At translation time, PyPy performs binding-time analysis of the source
RPython program after it has been turned to low-level graphs, i.e. at
the level at which operations manipulate pointer-and-structure-like
objects.

The binding-time terminology that we are using in PyPy is based on the
colors that we use when displaying the control flow graphs:

* *Green* variables contain values that are known at compile-time;
* *Red* variables contain values that are not known until run-time.

The binding-time analyzer of our translation tool-chain is based on the
same type inference engine that is used on the source RPython program,
the annotator.  In this mode, it is called the *hint-annotator*; it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.

The normal process of the hint-annotator is to propagate the binding
time (i.e. color) of the variables using the following kind of rules:

* For a foldable operation (i.e. one without side effect and which
  depends only on its argument values), if all arguments are green,
  then the result can be green too.

* Non-foldable operations always produce a red result.

* At join points, where multiple possible values (depending on control
  flow) are meeting into a fresh variable, if any incoming value comes
  from a red variable, the result is red.  Otherwise, the color of the
  result might be green.  We do not make it eagerly green, because of
  the control flow dependency: the residual function is basically a
  constant-folded copy of the source function, so it might retain some
  of the same control flow.  The value that needs to be stored in the
  fresh join variable thus depends on which branches are taken in the
  residual graph.

@@

The driving idea was that hints should be need-oriented.  Indeed, in a
program like an interpreter, there are a small number of places where it
would be clearly beneficial for a given value to be known at
compile-time, i.e. green: this is where we require the hints to be
added.

The hint-annotator assumes that all variables are red by default, and
then propagates annotations that record dependency information.
When encountering the user-provided hints, the dependency information
is used to make some variables green.  All
hints are in the form of an operation ``hint(v1, someflag=True)``
which semantically just returns its first argument unmodified.

The crucial need-oriented hint is ``v2 = hint(v1, concrete=True)``
which should be used in places where the programmer considers the
knowledge of the value to be essential.  This hint is interpreted by
the hint-annotator as a request for both ``v1`` and ``v2`` to be green.  It
has a *global* effect on the binding times: it means that not only
``v1`` but all the values that ``v1`` depends on -- recursively --
are forced to be green.  The hint-annotator complains if the
dependencies of ``v1`` include a value that cannot be green, like
a value read out of a field of a non-immutable structure.

Such a need-oriented backward propagation has advantages over the
commonly used forward propagation, in which a variable is compile-time
if and only if all the variables it depends on are also compile-time.  A
known issue with forward propagation is that it may mark as compile-time
either more variables than expected (which leads to over-specialization
of the residual code), or less variables than expected (preventing
specialization to occur where it would be the most useful).  Our
need-oriented approach reduces the problem of over-specialization, and
it prevents under-specialization: an unsatisfiable ``hint(v1,
concrete=True)`` is reported as an error.

In our context, though, such an error can be corrected.  This is done by
promoting a well-chosen variable among the ones that ``v1`` depends on.

Promotion is invoked with the use of a hint as well:
``v2 = hint(v1, promote=True)``.
This hint is a *local* request for ``v2`` to be green, without
requiring ``v1`` to be green.  Note that this amounts to copying
a red value into a green one, which is not possible in classical
approaches to partial evaluation.  See section \ref{promotion} for a
complete discussion of promotion.

For examples and further discussion on how the hints are applied in practice
see `Make your own JIT compiler` \cite{D08.1}.

@@

Once binding times (colors) have been assigned to all variables in a
family of control flow graphs, the next step is to mutate the graphs\footnote{
    One should keep in mind that the program described as the "source RPython
    program" in this document is typically an interpreter -- the canonical
    example is that it is the whole PyPy Standard Interpreter.  This
    program is meant to execute at run-time, and directly compute the
    intended result and side-effects. The translation process transforms
    it into a forest of flow graphs.  These are the flow graphs that
    timeshifting processes (and not the application-level program, which typically
    cannot be expressed as low-level flow graphs).
}
accordingly in order to produce a generating extension.  We call
this process *timeshifting* because it changes the time at
which the graphs are meant to be run, from run-time to compile-time.

@@

The basic idea of timeshifting is to transform operations in a way that
depends on the color of their operands and result.  Variables themselves
need to be represented based on their color:

* The red (run-time) variables have abstract values at compile-time;
  no actual value is available for them during compile-time. For them
  we use a boxed representation that can carry either a run-time storage
  location (a stack frame position or a register name) or an immediate
  constant (for when the value is, after all, known at compile-time).

* On the other hand, the green variables are the ones that can carry
  their value already at compile-time, so they are left untouched during
  timeshifting.

The operations of the original graphs are then transformed as follows:

* If an operation has no side effect nor any other run-time dependency, and
  if it only involves green operands, then it can stay unmodified in the
  graph.  In this case, the operation that was run-time in the original
  graph becomes a compile-time operation, and it will never be generated
  in the residual code.  (This is the case that makes the whole approach
  worthwhile: some operations become purely compile-time.)

* In all other cases, the operation might have to be generated in the
  residual code.  In the timeshifted graph it is replaced by a call
  to a helper which will generate a residual operation manipulating
  the input run-time values and return a new boxed representation
  for the run-time result location.

These helpers will constant-fold the operation if the inputs
are immediate constants and if the operation has no side-effects. Immediate constants can occur even though the
corresponding variable in the graph was red: a variable can be
dynamically found to contain a compile-time constant at a particular
point in (compile)-time, independently of the hint-annotator
proving that it is always the case.
In Partial Evaluation terminology, the timeshifted graphs are
performing some *on-line* partial evaluation in addition to the
off-line job enabled by the hint-annotator.

@@

After a *split,* i.e. after a conditional branch that could not be
folded at compile-time, the compilation state is duplicated and both
branches are compiled independently.  Conversely, after a *merge point,*
i.e. when two control flow paths meet each other, we try to join the two
paths in the residual code.  This part is more difficult because the two
paths may need to be compiled with different variable bindings -- e.g.
different variables may be known to take different compile-time constant
values in the two branches.  The two paths can either be kept separate
or merged; in the latter case, the merged compilation-time state needs
to be a generalization (*widening*) of the two already-seen states.
Deciding when to do each is a classical problem of partial evaluation,
as merging too eagerly may loose important precision and not merging
eagerly enough may create too many redundant residual code paths (to the
point of preventing termination of the compiler).

@@

In the sequel, we describe in more details one of the main new
techniques introduced in our approach, which we call *promotion*.  In
short, it allows an arbitrary run-time value to be turned into a
compile-time value at any point in time.  Each promotion point is
explicitly defined with a hint that must be put in the source code of
the interpreter.

From a partial evaluation point of view, promotion is the converse of
the operation generally known as "lift".  Lifting a value means
copying a variable whose binding time is compile-time into a variable
whose binding time is run-time -- it corresponds to the compiler
"forgetting" a particular value that it knew about.  By contrast,
promotion is a way for the compiler to gain *more* information about
the run-time execution of a program. Clearly, this requires
fine-grained feedback from run-time to compile-time, thus a
dynamic setting.

@@

A very different point of view on promotion is as a generalization of
techniques that already exist in dynamic compilers as found in modern
object-oriented language virtual machines.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{PIC}: the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results.

In the presence of promotion, dispatch optimization can usually be
reframed as a partial evaluation task.  Indeed, if the type of the
object being dispatched to is known at compile-time, the lookup can be
folded, and only a (possibly inlined) direct call remains in the
generated code.  In the case where the type of the object is not known
at compile-time, it can first be read at run-time out of the object and
promoted to compile-time.  As we will see in the sequel, this produces
very similar machine code.\footnote{
    This can also be seen as a generalization of a partial
    evaluation transformation called "The Trick" (see e.g. \cite{PE}),
    which again produces similar code but which is only
    applicable for finite sets of values.
}

@@

The implementation of promotion requires a tight coupling between
compile-time and run-time: a *callback,* put in the generated code,
which can invoke the compiler again.  When the callback is actually
reached at run-time, and only then, the compiler resumes and uses the
knowledge of the actual run-time value to generate more code.

@@

While this describes the general idea, the details are open to slight
variations.  Let us show more precisely the way the JIT compilers
produced by PyPy 1.0 work.  Our first example is purely artificial:

\begin{verbatim}
        ...
        b = a / 10
        c = hint(b, promote=True)
        d = c + 5
        print d
        ...
\end{verbatim}

In this example, ``a`` and ``b`` are run-time variables and ``c`` and
``d`` are compile-time variables; ``b`` is copied into ``c`` via a
promotion.  The division is a run-time operation while the addition is a
compile-time operation.

The compiler derived from an interpreter containing the above code
generates the following machine code (in pseudo-assembler notation),
assuming that ``a`` comes from register ``r1``:

\begin{verbatim}
     ...
        r2 = div r1, 10
     Label1:
        jump Label2
        <some reserved space here>

     Label2:
        call continue_compilation(r2, <state data pointer>)
        jump Label1
\end{verbatim}

The first time this machine code runs, the ``continue\_compilation()``
function resumes the compiler.  The two arguments to the function are
the actual run-time value from the register ``r2``, which the compiler
will now consider as a compile-time constant, and an immediate pointer
to data that was generated along with the above code snippet and which
contains enough information for the compiler to know where and with
which state it should resume.

Assuming that the first run-time value taken by ``r1`` is, say, 42, then
the compiler will see ``r2 == 4`` and update the above machine code as
follows:

\begin{verbatim}
     ...
        r2 = div r1, 10
     Label1:
        compare r2, 4            # patched
        jump-if-equal Label3     # patched
        jump Label2              # patched
        <less reserved space left>

     Label2:
        call continue_compilation(r2, <state data pointer>)
        jump Label1

@@

Notice how the addition is constant-folded by the compiler.  (Of course,
in real examples, different promoted values typically make the compiler
constant-fold complex code path choices in different ways, and not just
simple operations.)  Note also how the code following ``Label1`` is an
updatable switch which plays the role of a polymorphic inline cache.
The "polymorphic" terminology does not apply in our context, though, as
the switch does not necessarily have to be on the type of an object.

After the update, the original call to ``continue\_compilation()``
returns and execution loops back to the now-patched switch at
``Label1``.  This run and all following runs in which ``r1`` is between
40 and 49 will thus directly go to ``Label3``.  Obviously, if other
values show up, ``continue\_compilation()`` will be invoked again, so new
code will be generated and the code at ``Label1`` further patched to
check for more cases.

If, over the course of the execution of a program, too many cases are
seen, the reserved space after ``Label1`` will eventually run out.
Currently, we simply reserve more space elsewhere and patch the final
jump accordingly.  There could be better strategies which which we did
not implement so far, such as discarding old code and reusing their slots
in the switch, or sometimes giving up entirely and compiling a general
version of the code in which the value remains run-time.

@@

The *state data pointer* in the example above contains a snapshot of the
state of the compiler when it reached the promotion point.  Its memory
impact is potentially large -- a complete continuation for each generated
switch, which can never be reclaimed because new run-time values may
always show up later during the execution of the program.

To reduce the problem we compress the state into a so-called *path*.
The full state is only stored at a few specific points.\footnote{
    More precisely, at merge points that the user needs to mark
    as "global".  The control flow join point corresponding to the
    looping of the interpreter main loop is a typical place to put
    such a global merge point.
}
The compiler
records a trace of the multiple paths it followed from the last full
snapshot in a lightweight tree structure.  The *state data pointer* is
then only a pointer to a node in the tree; the branch from that node to
the root describes a path that let the compiler quickly *replay* its
actions (without generating code again) from the latest full snapshot to
rebuild its internal state and get back to the original promotion point.

For example, if the interpreter source code contains promotions inside a
run-time condition:

\begin{verbatim}
        if condition:
            ...
            hint(x, promote=True)
            ...
        else:
            ...
            hint(y, promote=True)
            ...
\end{verbatim}

>>>>

One of the central goals of the PyPy project \cite{PyPy} is to automatically
produce dynamic compilers from an interpreter, with as little
modifications of the interpreter code base itself as possible.

PyPy contains a complete interpreter for the Python language, written in
a high-level language, RPython, which is a subset of Python amenable to
static analysis.  It also contains a translation toolchain for compiling
this interpreter to either C (or C-like) environments, or to the higher
level environments provided by general-purpose virtual machines like
Java's and .NET.  The translation toolchain can input any RPython
program, although our focus was on the translation of RPython programs
that are interpreters for dynamic languages.\footnote{We also have an
interpreter for Prolog and the beginning of one for JavaScript.}

The translation framework uses control flow graphs in SSI format as its
intermediate representation (SSI is a stricter subset of SSA).  The
details of this process are beyond the scope of the present paper, and
have been presented in \cite{pypyvmconstruction}.
The present paper describes a
special optional transformation that we integrated with this translation
framework: deriving a dynamic compiler from the interpreter.  In other
words, our translation framework is able to input an interpreter for any
language (it works best for dynamic languages); as long as it is
written in RPython and contains a small number of extra hints, then it
can produce from it a complete virtual machine
\emph{that contains a just-in-time compiler for the dynamic language.}

Partial evaluation techniques should, at least theoretically,
allow such a derivation of a compiler from an interpreter
\cite{partial-evaluation}, but it
is not reasonable to expect the code produced for an input program by
a compiler derived using partial evaluation to be very good,
especially in the case of a dynamic language.  Essentially, the input
program doesn't contain enough information to generate good code; for
example the input program contains mostly no kind of type
information in that case.

@@

Inspired by Psyco \cite{psyco-paper}, which is a hand-written dynamic compiler
based on partial evaluation for Python, we developed a technique -
\emph{promotion} - for our dynamic compiler generator. Simply put, promotion
on a value stops compilation and waits until the run-time reaches this
point.  When it does, the actual run-time value is promoted into a
compile-time constant, and compilation resumes with this extra
information.

@@

\def\code#1{\texttt{#1}}

Partial evaluation is the process of evaluating a function, say \code{f(x,
y)}, with only partial information about the values of its arguments,
say the value of the \code{x} argument only.  This produces a \emph{residual}
function \code{g(y)}, which takes less arguments than the original -- only
the information not specified during the partial evaluation process needs
to be provided to the residual function, in this example the \code{y}
argument.

Partial evaluation (PE) comes in two flavors:
%
\begin{enumerate}

\item\emph{On-line PE:} a compiler-like algorithm takes the source code of the
  function \code{f(x, y)} (or its intermediate representation, i.e.\ its
  control flow graph in PyPy's terminology), and some partial
  information, e.g.\ \code{x=5}.  From this, it produces the residual
  function \code{g(y)} directly, by following in which operations the
  knowledge \code{x=5} can be used, which loops can be unrolled, etc.

\item\emph{Off-line PE:} in many cases, the goal of partial evaluation is to
  improve performance in a specific application.  Assume that we have a
  single known function \code{f(x, y)} in which we think that the value of
  \code{x} will change slowly during the execution of our program -- much
  more slowly than the value of \code{y}.  An obvious example is a loop
  that calls \code{f(x, y)} many times with always the same value \code{x}.
  We could then use an on-line partial evaluator to produce a \code{g(y)}
  for each new value of \code{x}.  In practice, the overhead of the partial
  evaluator might be too large for it to be executed at run-time.
  However, if we know the function \code{f} in advance, and if we know
  \emph{which} arguments are the ones that we will want to partially evaluate
  \code{f} with, then we do not need a full compiler-like analysis of \code{f}
  every time the value of \code{x} changes.  We can precompute once and for
  all a specialized function \code{f1(x)}, which when called produces the
  residual function \code{g(y)} corresponding to \code{x}.  This is
  \emph{off-line partial evaluation;} the specialized function \code{f1(x)}
  is called a \emph{generating extension.}

\end{enumerate}

The PyPy JIT generation framework is based on off-line partial
evaluation.  The function called \code{f(x, y)} above is typically the main
loop of some interpreter written in RPython.  The size of the interpreter can range
from a three-liner used for testing purposes to the whole of PyPy's
Python interpreter.  In all cases, \code{x} stands for the input program
(the bytecode to interpret) and \code{y} stands for the input data (like a
frame object with the binding of the input arguments and local
variables).  Our framework is capable of automatically producing the
corresponding generating extension \code{f1(x)}, which takes an input
program only and produces a residual function \code{g(y)}.  This \code{f1(x)}
is a compiler\footnote{
    What we get in PyPy is more precisely a \emph{just-in-time compiler:}
    if promotion is used, compiling ahead of time is not possible.
}
for the very same language for which \code{f(x, y)} is
an interpreter.

Off-line partial evaluation is based on \emph{binding-time analysis,} which
is the process of determining among the variables used in a function (or
a set of functions) which ones are going to be known in advance and
which ones are not.  In the example of \code{f(x, y)}, such an analysis
would be able to infer that the constantness of the argument \code{x}
implies the constantness of many intermediate values used in the
function.  The \emph{binding time} of a variable determines how early the
value of the variable will be known.

Once binding times have been determined, one possible approach to
producing the generating extension itself is by self-applying on-line
partial evaluators.  This is known as the second Futamura projection
\cite{Futamura}.  So far it is unclear if this approach can lead to optimal
results, or even if it scales well.  In PyPy we selected a more direct
approach: the generating extension is produced by transformation of the
control flow graphs of the interpreter, guided by the binding times.  We
call this process \emph{timeshifting.}

\subsection{Related work}

XXX PE; Psyco; REJIT; ?

@@

PyPy contains a framework for generating just-in-time compilers using
off-line partial evaluation.  As such, there are three distinct phases:
%
\begin{enumerate}

\item\emph{Translation time:} during the normal translation of an RPython
  program, say PyPy's Python interpreter, we perform binding-time
  analysis and off-line specialization ("timeshifting") of the
  interpreter.  This produces a generating extension, which is linked
  with the rest of the program.

\item\emph{Compile time:} during the execution of the program, when a new
  bytecode is about to be interpreted, the generating extension is
  invoked instead.  As the generating extension is a compiler, all the
  computations it performs are called compile-time computations.  Its
  sole effect is to produce residual code.

\item\emph{Run time:} the normal execution of the program (which includes the
  time spent running the residual code created by the generating
  extension).

\end{enumerate}

@@

At translation time, PyPy performs binding-time analysis of the source
RPython program after it has been turned to low-level graphs, i.e.\ at
the level at which operations manipulate pointer-and-structure-like
objects.

The binding-time terminology that we are using in PyPy is based on the
colors that we use when displaying the control flow graphs:
%
\begin{itemize}
\item\emph{Green} variables contain values that are known at compile-time;
\item\emph{Red} variables contain values that are not known until run-time.
\end{itemize}

The binding-time analyzer of our translation tool-chain is based on the
same type inference engine that is used on the source RPython program,
the annotator.  In this mode, it is called the \emph{hint-annotator;} it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.

The normal process of the hint-annotator is to propagate the binding
time (i.e.\ color) of the variables using the following kind of rules:
%
\begin{itemize}

\item For a foldable operation (i.e.\ one without side effect and which
  depends only on its argument values), if all arguments are green,
  then the result can be green too.

\item Non-foldable operations always produce a red result.

\item At join points, where multiple possible values (depending on control
  flow) are meeting into a fresh variable, if any incoming value comes
  from a red variable, the result is red.  Otherwise, the color of the
  result might be green.  We do not make it eagerly green, because of
  the control flow dependency: the residual function is basically a
  constant-folded copy of the source function, so it might retain some
  of the same control flow.  The value that needs to be stored in the
  fresh join variable thus depends on which branches are taken in the
  residual graph.

\end{itemize}

@@

The driving idea was that hints should be need-oriented.  Indeed, in a
program like an interpreter, there are a small number of places where it
would be clearly beneficial for a given value to be known at
compile-time, i.e.\ green: this is where we require the hints to be
added.

The hint-annotator assumes that all variables are red by default, and
then propagates annotations that record dependency information.
When encountering the user-provided hints, the dependency information
is used to make some variables green.  All
hints are in the form of an operation \code{hint(v1, someflag=True)}
which semantically just returns its first argument unmodified.

The crucial need-oriented hint is
$$\code{v2 = hint(v1, concrete=True)}$$
which should be used in places where the programmer considers the
knowledge of the value to be essential.  This hint is interpreted by
the hint-annotator as a request for both \code{v1} and \code{v2} to be green.  It
has a \emph{global} effect on the binding times: it means that not only
\code{v1} but all the values that \code{v1} depends on -- recursively --
are forced to be green.  The hint-annotator complains if the
dependencies of \code{v1} include a value that cannot be green, like
a value read out of a field of a non-immutable structure.

Such a need-oriented backward propagation has advantages over the
commonly used forward propagation, in which a variable is compile-time
if and only if all the variables it depends on are also compile-time.  A
known issue with forward propagation is that it may mark as compile-time
either more variables than expected (which leads to over-specialization
of the residual code), or less variables than expected (preventing
specialization to occur where it would be the most useful).  Our
need-oriented approach reduces the problem of over-specialization, and
it prevents under-specialization: an unsatisfiable \code{hint(v1,
concrete=True)} is reported as an error.

In our context, though, such an error can be corrected.  This is done by
promoting a well-chosen variable among the ones that \code{v1} depends on.

Promotion is invoked with the use of a hint as well:
\code{v2 = hint(v1, promote=True)}.
This hint is a \emph{local} request for \code{v2} to be green, without
requiring \code{v1} to be green.  Note that this amounts to copying
a red value into a green one, which is not possible in classical
approaches to partial evaluation.  See section \ref{promotion} for a
complete discussion of promotion.

For examples and further discussion on how the hints are applied in practice
see \emph{Make your own JIT compiler} at
\code{http://codespeak.net/pypy/dist/pypy/doc/jit.html}. % XXX check url

@@

Once binding times (colors) have been assigned to all variables in a
family of control flow graphs, the next step is to mutate the graphs\footnote{
    One should keep in mind that the program described as the "source RPython
    program" in this document is typically an interpreter -- the canonical
    example is that it is the whole PyPy Standard Interpreter.  This
    program is meant to execute at run-time, and directly compute the
    intended result and side-effects. The translation process transforms
    it into a forest of flow graphs.  These are the flow graphs that
    timeshifting processes (and not the application-level program, which typically
    cannot be expressed as low-level flow graphs).
}
accordingly in order to produce a generating extension.  We call
this process \emph{timeshifting} because it changes the time at
which the graphs are meant to be run, from run-time to compile-time.

@@

The basic idea of timeshifting is to transform operations in a way that
depends on the color of their operands and result.  Variables themselves
need to be represented based on their color:
%
\begin{itemize}

\item The red (run-time) variables have abstract values at compile-time;
  no actual value is available for them during compile-time. For them
  we use a boxed representation that can carry either a run-time storage
  location (a stack frame position or a register name) or an immediate
  constant (for when the value is, after all, known at compile-time).

\item On the other hand, the green variables are the ones that can carry
  their value already at compile-time, so they are left untouched during
  timeshifting.

\end{itemize}

The operations of the original graphs are then transformed as follows:
%
\begin{itemize}

\item If an operation has no side effect nor any other run-time dependency, and
  if it only involves green operands, then it can stay unmodified in the
  graph.  In this case, the operation that was run-time in the original
  graph becomes a compile-time operation, and it will never be generated
  in the residual code.  (This is the case that makes the whole approach
  worthwhile: some operations become purely compile-time.)

\item In all other cases, the operation might have to be generated in the
  residual code.  In the timeshifted graph it is replaced by a call
  to a helper which will generate a residual operation manipulating
  the input run-time values and return a new boxed representation
  for the run-time result location.

\end{itemize}

These helpers will constant-fold the operation if the inputs
are immediate constants and if the operation has no side-effects. Immediate constants can occur even though the
corresponding variable in the graph was red: a variable can be
dynamically found to contain a compile-time constant at a particular
point in (compile)-time, independently of the hint-annotator
proving that it is always the case.
In Partial Evaluation terminology, the timeshifted graphs are
performing some \emph{on-line} partial evaluation in addition to the
off-line job enabled by the hint-annotator.

@@

After a \emph{split,} i.e.\ after a conditional branch that could not be
folded at compile-time, the compilation state is duplicated and both
branches are compiled independently.  Conversely, after a \emph{merge point,}
i.e.\ when two control flow paths meet each other, we try to join the two
paths in the residual code.  This part is more difficult because the two
paths may need to be compiled with different variable bindings --
e.g.\ different variables may be known to take different compile-time constant
values in the two branches.  The two paths can either be kept separate
or merged; in the latter case, the merged compilation-time state needs
to be a generalization \emph{(widening)} of the two already-seen states.
Deciding when to do each is a classical problem of partial evaluation,
as merging too eagerly may loose important precision and not merging
eagerly enough may create too many redundant residual code paths (to the
point of preventing termination of the compiler).

@@

In the sequel, we describe in more details one of the main new
techniques introduced in our approach, which we call \emph{promotion.}  In
short, it allows an arbitrary run-time value to be turned into a
compile-time value at any point in time.  Each promotion point is
explicitly defined with a hint that must be put in the source code of
the interpreter.

From a partial evaluation point of view, promotion is the converse of
the operation generally known as "lift".  Lifting a value means
copying a variable whose binding time is compile-time into a variable
whose binding time is run-time -- it corresponds to the compiler
"forgetting" a particular value that it knew about.  By contrast,
promotion is a way for the compiler to gain \emph{more} information about
the run-time execution of a program. Clearly, this requires
fine-grained feedback from run-time to compile-time, thus a
dynamic setting.

@@

A very different point of view on promotion is as a generalization of
techniques that already exist in dynamic compilers as found in modern
object-oriented language virtual machines.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{polymorphic-inline-caches}:
the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results.

In the presence of promotion, dispatch optimization can usually be
reframed as a partial evaluation task.  Indeed, if the type of the
object being dispatched to is known at compile-time, the lookup can be
folded, and only a (possibly inlined) direct call remains in the
generated code.  In the case where the type of the object is not known
at compile-time, it can first be read at run-time out of the object and
promoted to compile-time.  As we will see in the sequel, this produces
very similar machine code.\footnote{
    This can also be seen as a generalization of a partial
    evaluation transformation called "The Trick" (see e.g.\ \cite{partial-evaluation}),
    which again produces similar code but which is only
    applicable for finite sets of values.
}

@@

The implementation of promotion requires a tight coupling between
compile-time and run-time: a \emph{callback,} put in the generated code,
which can invoke the compiler again.  When the callback is actually
reached at run-time, and only then, the compiler resumes and uses the
knowledge of the actual run-time value to generate more code.

@@

While this describes the general idea, the details are open to slight
variations.  Let us show more precisely the way the JIT compilers
produced by PyPy 1.0 work.  Our first example is purely artificial:
%
\begin{verbatim}
    ...
    b = a / 10
    c = hint(b, promote=True)
    d = c + 5
    print d
    ...
\end{verbatim}

In this example, \code{a} and \code{b} are run-time variables and \code{c} and
\code{d} are compile-time variables; \code{b} is copied into \code{c} via a
promotion.  The division is a run-time operation while the addition is a
compile-time operation.

The compiler derived from an interpreter containing the above code
generates the following machine code (in pseudo-assembler notation),
assuming that \code{a} comes from register \code{r1}:
%
\begin{verbatim}
 ...
    r2 = div r1, 10
 Label1:
    jump Label2
    <some reserved space here>

 Label2:
    call continue_compilation(r2, <state data ptr>)
    jump Label1
\end{verbatim}

The first time this machine code runs, the function called
\code{continue\_compilation()}
resumes the compiler.  The two arguments to the function are
the actual run-time value from the register \code{r2}, which the compiler
will now consider as a compile-time constant, and an immediate pointer
to data that was generated along with the above code snippet and which
contains enough information for the compiler to know where and with
which state it should resume.

Assuming that the first run-time value taken by \code{r1} is, say, 42, then
the compiler will see \code{r2 == 4} and update the above machine code as
follows:
%
\begin{verbatim}
 ...
    r2 = div r1, 10
 Label1:
    compare r2, 4            # patched
    jump-if-equal Label3     # patched
    jump Label2              # patched
    <less reserved space left>

 Label2:
    call continue_compilation(r2, <state data ptr>)
    jump Label1

@@

Notice how the addition is constant-folded by the compiler.  (Of course,
in real examples, different promoted values typically make the compiler
constant-fold complex code path choices in different ways, and not just
simple operations.)  Note also how the code following \code{Label1} is an
updatable switch which plays the role of a polymorphic inline cache.
The "polymorphic" terminology does not apply in our context, though, as
the switch does not necessarily have to be on the type of an object.

After the update, the original call to \code{continue\_compilation()}
returns and execution loops back to the now-patched switch at
\code{Label1}.  This run and all following runs in which \code{r1} is between
40 and 49 will thus directly go to \code{Label3}.  Obviously, if other
values show up, \code{continue\_compilation()} will be invoked again, so new
code will be generated and the code at \code{Label1} further patched to
check for more cases.

If, over the course of the execution of a program, too many cases are
seen, the reserved space after \code{Label1} will eventually run out.
Currently, we simply reserve more space elsewhere and patch the final
jump accordingly.  There could be better strategies which which we did
not implement so far, such as discarding old code and reusing their slots
in the switch, or sometimes giving up entirely and compiling a general
version of the code in which the value remains run-time.

@@

The state data pointer in the example above contains a snapshot of the
state of the compiler when it reached the promotion point.  Its memory
impact is potentially large -- a complete continuation for each generated
switch, which can never be reclaimed because new run-time values may
always show up later during the execution of the program.

To reduce the problem we compress the state into a so-called \emph{path.}
The full state is only stored at a few specific points.\footnote{
    More precisely, at merge points that the user needs to mark
    as "global".  The control flow join point corresponding to the
    looping of the interpreter main loop is a typical place to put
    such a global merge point.
}
The compiler
records a trace of the multiple paths it followed from the last full
snapshot in a lightweight tree structure.  The state data pointer is
then only a pointer to a node in the tree; the branch from that node to
the root describes a path that let the compiler quickly \emph{replay} its
actions (without generating code again) from the latest full snapshot to
rebuild its internal state and get back to the original promotion point.

For example, if the interpreter source code contains promotions inside a
run-time condition:
%
\begin{verbatim}
        if condition:
            ...
            hint(x, promote=True)
            ...
        else:
            ...
            hint(y, promote=True)
            ...
\end{verbatim}
